Append-only Datastore

### 2020-06-01 paper/paper.tex
Finished first draft of the follower

\begin{enumerate}
	\item Clients communicate only with their nearest follower for extremely
		low latency.
	\item Followers provide linearizability for client appends on a single
		key. That is, all clients writing to the same key of the same
		follower will see linearized writes.
	\item Eventual consistency is minimally disruptive to clients. A client
		may read the data for key $k$ and receive data consisting of 2
		writes: $w_1, w_2$. Write $w_x$ sent through another follower
		preserves the ordering of $w_1$ and $w_2$, so the system
		eventually returns $w_x, w_1, w_2$, $w_1, w_x, w_2$, or $w_1,
		w_2, w_x$ when read.  Because of (2), writes directly to the
		same follower are linearizable.
\end{enumerate}

@@

Clients are exposed the following API by followers:
\begin{itemize}
	\item \texttt{append(key, data)} - Appends \texttt{data} to the existing
		data for \texttt{key}. Guarantees that \texttt{data} appears
		after any data that the client already read for \texttt{key}.
	\item \texttt{get(key)} - Gets the data for \texttt{key}, expressed as a
		list of indexed values. For example, the returned data will be
		the result of writes 1, 3, and 5, expressed as $\{w_1, w_3,
		w_5\}$. Subsequent \texttt{get} requests will always return
		$w_1$, $w_3$, and $w_5$ in the same relative order, but other
		writes may be prepended, appended, or interspersed.
\end{itemize}

@@

\texttt{append} requests are also fast, limited by the speed of a local disk
write. When a client appends data to key \texttt{k}, the follower writes that
data - along with its index - to local disk before returning to the client. This
ensures that data is not lost upon crashing and rebooting. For increased fault
tolerance, the append should be written to multiple disks locally or to a remote
disk.

@@

There are two types of responses. The first - and likely most common - case
is a simple \texttt{SUCCESS} response. This indicates that the append committed
successfully and the follower has up-to-date data for that key. The second is a
\texttt{NEEDS\_UPDATE} response. This indicates that the data was appended, but
another \texttt{update} preempted it. In this scenario, the follower changes the
locally stored index of the appended data, and sends a \texttt{sync} request
another follower indicated by the leader's response.

@@

In the common case, where updates do not overlap...

When concurrent updates do occur...

Both \texttt{update} and \texttt{sync} requests are retried in case of network
failure. If a follower reboots during a \texttt{update}, it can inspect the disk
at boot and see that it has appended data locally that has not been confirmed by
the leader. If a follower reboots during a \texttt{sync}, it will have stale
data until a \texttt{get} or \texttt{append} causes it to update. This keeps the
system simple while retaining its consistency properties.

@@

\subsection{Common Case}

\subsection{Corner Cases}
% Multiple simultaneous writes from c1 and c2 to f1. How do we ensure that the
% GL does not order w2 before w1?

>>>>

\begin{enumerate}
	\item Clients communicate only with their nearest follower for extremely
		low latency.
	\item Followers provide linearizability for client appends on a single
		key. That is, all clients writing to the same key of the same
		follower will see linearized writes.
	\item Eventual consistency is minimally disruptive to clients. A client
		may read the data for key $k$ and receive data consisting of 2
		writes' data: $d_1, d_2$. Write $x$ with data $d_x$ sent through another follower
		preserves the ordering of $d_1$ and $d_2$, so the system
		eventually returns $d_x, d_1, d_2$, $d_1, d_x, d_2$, or $d_1,
		d_2, d_x$ when read.  Because of (2), writes directly to the
		same follower are linearizable.
\end{enumerate}

@@

Clients are exposed the following API by followers:
\begin{itemize}
	\item \texttt{append(key, data)} - Appends \texttt{data} to the existing
		data for \texttt{key}. Guarantees that \texttt{data} appears
		after any data that the client already read for \texttt{key}.
	\item \texttt{get(key)} - Gets the data for \texttt{key}, expressed as a
		list of indexed values. For example, the returned data will be
		the result of writes 1, 3, and 5, expressed as $\{d_1, d_3,
		d_5\}$. Subsequent \texttt{get} requests will always return
		$d_1$, $d_3$, and $d_5$ in the same relative order, but other
		writes may be prepended, appended, or interspersed.
\end{itemize}

@@

\texttt{append} requests are also fast, limited by the speed of a local disk
write. When a client appends data to key \texttt{k}, the follower writes that
data - along with its index - to local disk before returning to the client. This
ensures that data is not lost upon crashing and rebooting. For increased fault
tolerance, the append should be written to multiple disks locally or to a remote
disk. Note that throughout this section when we refer to updating local storage,
the in-memory store is also updated immediately after data is written to disk.

@@

There are two types of responses. The first - and likely most common - case
is a simple \texttt{SUCCESS} response. This indicates that the append completed
successfully and the follower has up-to-date data for that key. The second is a
\texttt{NEEDS\_UPDATE} response. This indicates that the data was appended, but
another \texttt{update} preempted it. In this scenario, the follower changes the
locally stored index of the appended data, and sends a \texttt{sync} request
another follower indicated by the leader's response.

@@

\subsubsection{Common Case}
When updates do not overlap, the follower handles \texttt{append} requests as
follows:

\begin{enumerate}
	\item Write the append to local storage along with the index \texttt{i}
		the follower believes applies to the append.
	\item Respond to the client that the request was successful.
	\item Concurrently with (2), send an \texttt{update} to the leader that
		an append occured on key \texttt{k} at index \texttt{i}.
	\item The leader responds with \texttt{SUCCESS}.
	\item The follower updates local disk to indicate that the append has
		been commited at index \texttt{i} by the leader.
\end{enumerate}

The data for key \texttt{k} has simply benn appended to, changing from $\{d_a,
d_b, d_c, ..., d_{i-1}\}$ to $\{d_a, d_b, d_c, ..., d_{i-1}, d_i\}$, where $a < b <
c < i$. Note that there may be data ``holes" between $a$, $b$, $c$, and $i$ not
yet filled in by either a \texttt{get} or \texttt{sync}.

\subsubsection{Conccurent Updates}
When concurrent updates occur:

% TODO: There's some ugly mixing of texttt and math notation here, particularly
% around i'.
\begin{enumerate}
	\item The same as steps 1-3 above.
	\item The leader responds with \texttt{NEEDS\_UPDATE}.
	\item The follower updates local disk to indicate that the append has
		been commited at index \texttt{i`} by the leader.
	\item The follower sends a \texttt{sync} request to the follower that
		appended the data at index \texttt{i`-1}, as indicated in the
		leader's response.
	\item The follower receives and stores on local disk the data it was
		missing, including what the leader deemed to be the ``winner" of
		index \texttt{i}.
\end{enumerate}

The data for key \texttt{k} in this case is an append and a reordering,
changing from $\{d_a, d_b, d_c, ..., d_{i-1}\}$ to $\{d_a, d_b, d_c, ...,
d_{i-1}, d_i, d_{i'}\}$, where $a < b < c < i$ and as above may have more holes
to fill.

% TODO: We don't yet implement the early return of data.
It is possible for a chain of \texttt{sync} requests to form when more than 2
followers append a key at the same time. If follower $a$ starts syncing with
follower $b$, which is already waiting on follower $c$ to respond to a
\texttt{sync}, $b$ will send what is knows to $a$. $a$ then retries to get
missing data, at which point $b$ waits to respond until it has new data to send.

\subsubsection{Follower Fault Tolerance}
Both \texttt{update} and \texttt{sync} requests are retried in case of network
failure. If a follower reboots during a \texttt{update}, it can inspect the disk
at boot and see that it has appended data locally that has not been confirmed by
the leader. If a follower reboots during a \texttt{sync}, it will have stale
data until a \texttt{get} or \texttt{append} causes it to update. This keeps the
system simple while retaining its consistency properties.

@@

\subsection{Corner Cases}
% Multiple simultaneous writes from c1 and c2 to f1. How do we ensure that the
% GL does not order w2 before w1?
% What if the sync request can be partially satisfied. Do we not respond, or
% respond partially and expect another sync request from the same follower?
